Using Hybrid Machine Learning for Fundamental Stellar Parameter Inference

### 2024-04-19 paper/paper2_neo/paper.tex
Add figures and math to paper

Multiple efforts have been made in this direction, treating spectra in 
different ways. The standard practice is to treat the wavelength and flux as 
simply two arrays and implement any from a myriad of different algorithms to 
inference fundamental stellar parameters. Others take a more physics-informed 
approach by decomposing spectra into some eigenbasis and treating it as a 
regression problem, such as \texttt{starfish}. However, \texttt{blase} takes 
a different approach. It treats spectra not as a set of pixels or a set of 
eigenbasis coefficients but as a set of spectral lines, specifically Voigt 
profiles. There is no metric of this approach being better or worse than the 
others mentioned so far, but it is a different paradigm and is worth 
exploring as a semi-empirical tool. This study aims to take advantage of 
\texttt{blase}'s ability to change a spectrum's basis from flux versus 
wavelength into spectral line versus spectral line properties, and use that 
to develop a robust tool for stellar parameter inference.

@@

\begin{table}[h]
    \hspace*{-1.35cm}\begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\textbf{PHOENIX Grid Subset}}\\
        \hline\hline
        Parameter & Symbol & Range\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0] dex\\
        Iron Abundance & [Fe/H] & [-0.5, 0] dex\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K\\
        Surface Gravity & $\log(g)$ & [2, 6]\\
        Wavelength & $\lambda$ & [8038, 12849] \AA\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study. These limits 
    were imposed to reduce the computational cost of the algorithms and to 
    ensure a rectilinear parameter space in order to work with \texttt{scipy}'s
    \texttt{RegularGridInterpolator}. The wavelength limits in particular 
    roughly line up with that of the Habitable Zone Planet Finder (HPF) 
    spectrograph.}
\end{table}

\subsection{Preprocessing with \texttt{gollum}}
First, the PHOENIX subset was accessed directly from the PHOENIX website
using \texttt{gollum}'s download option. The spectra were then put through a 
three-step preprocessing pipeline.
\begin{enumerate}
    \item Blackbody Division: Since the $T_{\mathrm{eff}}$ of each spectrum 
    is known, the according blackbody spectrum was divided out.
    \item Percentile Normalization: The spectra were normalized by dividing
    them by their 99th percentile.
    \item Continuum Normalization: The spectra were further normalized by 
    dividing them by a \nth{5} order polynomial continuum.
\end{enumerate}

@@

Once optimization was complete, the list of identified lines, present in the
state dictionary of \texttt{blase}'s model, was saved to a \texttt{.pt} file
for each PHOENIX subset grid point.

\section{Interpolating Manifolds}
\subsection{Cross-Model Line Identification}
Earlier we mentioned that \texttt{blase} tuned the line centers of
detected lines. This means that from one PHOENIX spectrum to the next, the
same line could have a slightly different line center. Since the goal of this
study is to interpolate the properties of each line, we needed to identify the
presence of a particular line across the PHOENIX subset. We decided to do
this by using the line centers of the detected lines pre-optimization. This
introduced a new variable called the jitter $\delta\mu$, which measures the shift of the
line center between pre-optimization and post-optimization. Now with each
spectral line indexed by $\mu$, we had four parameters to interpolate:
$\delta\mu$, $\ln(a)$, $\sigma$, and $\gamma$.

There was also a second issue, and that is that spectral lines were often
only detected in some spectra from the PHOENIX subset. This means that in 
theory, interpolation would be inaccurate in regions where a line does not 
appear, and also breaks in practice because \texttt{scipy}'s regular grid
interpolation relies on a rectilinear grid, which we would not have if
regions of grid points were missing. To solve this, we artificially
populated missing grid points with log-amplitudes of -1000, which retained
interpolator stability by not being an infinity, but also essentially 
nullified the line when evaluated.

\subsection{Continuously Evaluable Manifolds}
For each line, the inputs to the interpolator were the three input parameters $T_{\mathrm{eff}}$,
$\log(g)$, and [Fe/H], and the output was a list of four parameters, 
$\delta\mu$, $\ln(a)$, $\sigma$, and $\gamma$. For each line, one of these 
interpolator objects was created using linear interpolation, and these 
interpolators were aggregated into a single list, which was when pickled 
and written to a \texttt{.pkl} file. These interpolators represent multiple 
manifolds mapping fundamental stellar parameters to spectral line properties. 
These interpolators could now be evaluated at any point lying within the 
domain of the PHOENIX subset, turning a discretely sampled PHOENIX subset 
into a continuous one. With the given size of the PHOENIX subset, the 
interpolator list takes up 13.2 GB of disk space. Since this is now evaluable,
we call this the PHOENIX generator.

\section{Bayesian Inference and Testing}
\subsection{Inference Algorithm}
With the PHOENIX generator complete, we used Bayesian optimization as the
inference algorithm, specifically the \texttt{gp\_minimize} function from
the \texttt{scikit-optimize} library. This algorithm uses a Gaussian Process
to model the objective function, which in this case was the RMS 
(Root-Mean-Square) loss between the interpolated spectrum and the true 
spectrum. The optimizer was configured to first run 30 random evaluations to 
seed the surrogate model, then run 20 more evaluations now guided by the 
surrogate model. This totals to 50 evaluations, which was deemed sufficient 
for this study. We then decided to use ensemble methods, running 32 parallel
optimizers to get a more statistically sound result. One inference run 
takes an average of [TIME] to complete on Triton.

@@

\section{Discussion}
The computer used for this study, Triton, has the following specifications:
\begin{table}[h]
    \hspace*{0.6cm}\begin{tabular}{|c|c|}
        \hline
        \multicolumn{2}{|c|}{\textbf{Triton Specs}}\\
        \hline\hline
        CPU & AMD EPYC 7513\\
        Memory & 256 GB\\
        GPU & Nvidia A100 40GB ($\times$2)\\
        \hline
    \end{tabular}
    \caption{This computer is owned by Caroline Morley and is available to
    members of her research group. It was used for all computations, but not 
    for generating visualizations. The EPYC 7513 is a 32c/64t CPU with a 
    boost clock of 3.65 GHz.}
\end{table}

>>>>

Multiple efforts have been made in this direction, treating spectra in 
different ways. The standard practice is to treat the wavelength and flux as 
simply two arrays and implement any from a myriad of different algorithms to 
inference fundamental stellar parameters. Others take a more mathematical 
approach by decomposing spectra into some eigenbasis and treating it as a 
regression problem, such as \texttt{starfish}. However, \texttt{blase} takes 
a different approach. It treats spectra not as a set of pixels or a set of 
eigenbasis coefficients but as a set of spectral lines, specifically Voigt 
profiles. There is no metric of this approach being better or worse than the 
others mentioned so far, but it is a different paradigm and is worth 
exploring as a semi-empirical tool. This study aims to take advantage of 
\texttt{blase}'s ability to change a spectrum's basis from flux versus 
wavelength into spectral line versus spectral line properties, and use that 
to develop a robust tool for stellar parameter inference.

@@

\begin{table}[h!]
    \hspace*{-1.35cm}\begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\textbf{PHOENIX Grid Subset}}\\
        \hline\hline
        Parameter & Symbol & Range\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0] dex\\
        Iron Abundance & [Fe/H] & [-0.5, 0] dex\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K\\
        Surface Gravity & $\log(g)$ & [2, 6]\\
        Wavelength & $\lambda$ & [8038, 12849] \AA\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study. These limits 
    were imposed to reduce the computational cost of the algorithms and to 
    ensure a rectilinear parameter space in order to work with \texttt{scipy}'s
    \texttt{RegularGridInterpolator}. The wavelength limits in particular 
    roughly line up with that of the Habitable Zone Planet Finder (HPF) 
    spectrograph. This subset is comprised of 1314 individual spectra.}
\end{table}

\subsection{Preprocessing with \texttt{gollum}}
First, the PHOENIX subset was accessed directly from the PHOENIX website
using \texttt{gollum}'s download option. The spectra were then put through a 
three-step preprocessing pipeline.
\begin{enumerate}
    \item \textit{Blackbody Division}: Since the $T_{\mathrm{eff}}$ of each spectrum 
    is known, the according blackbody spectrum was divided out.
    \item \textit{Percentile Normalization}: The spectra were normalized by dividing
    them by their 99th percentile.
    \item \textit{Continuum Normalization}: The spectra were further normalized by 
    dividing them by a \nth{5} order polynomial continuum.
\end{enumerate}
Mathematically, we can express the preprocessing as follows:
\begin{gather}
    \mathsf{\bar{S}} = \dfrac{\mathsf{S}}{\mathsf{B}\mathsf{Q}_5P_{99}}
\end{gather}
where $\mathsf{\bar{S}}$ is the preprocessed spectrum, $\mathsf{S}$ is the 
original spectrum,  $\mathsf{B}$ is the blackbody spectrum, $\mathsf{Q}_n$ 
is the $n^\mathrm{th}$ order polynomial continuum fit, and $P_n$ is the 
$n^\mathrm{th}$ percentile function. Arithmetic operations between arrays 
are assumed to be elementwise in all notation from here on out.

@@

Once optimization was complete, the list of identified lines, present in the
state dictionary of \texttt{blase}'s model, was saved to a \texttt{.pt} file
for each PHOENIX subset grid point. The total disk space these files took up 
is 465 MB.

\section{Interpolating Manifolds}
\subsection{Cross-Model Line Identification}
Earlier we mentioned that \texttt{blase} tuned the line centers of
detected lines. This means that from one PHOENIX spectrum to the next, the
same line could have a slightly different line center. Since the goal of this
study is to interpolate the properties of each line, we needed to identify the
presence of a particular line across the PHOENIX subset. We decided to do
this by using the line centers $\mu$ of the detected lines pre-optimization, 
while the post-optimization line centers are labeled $\mu'$. Now with 
each spectral line indexed by $\mu$, we had four parameters to interpolate:
$\mu'$, $\ln(a)$, $\sigma$, and $\gamma$. Mathematically, the Voigt profile
$\mathsf{V}_\mu$ in terms of our four parameters is given by:

\begin{gather}
    \mathsf{V}_\mu(\lambda) = e^{\ln(a)}\mathrm{Re}\left[w\left(\dfrac{\lambda - \mu' + i\gamma}{\sigma\sqrt{2}}\right)\right] \bigg/ \left(\sigma\sqrt{2\pi}\right)
\end{gather}
where $w$ is the Faddeeva function. Note that since we are dealing with 
the parameters of these Voigt profiles, even if the interpolation method is 
linear, the actual result of interpolation varies non-linearly.

There is also a second issue, and that is that spectral lines were often
only detected in some spectra from the PHOENIX subset. In Figure 1, we show
that different grid points sport differing counts of detected spectral lines.
This means that in theory, interpolation would be inaccurate in regions where 
a line does not appear, and also breaks in practice because \texttt{scipy}'s 
\texttt{RegularGridInterpolator} relies on a rectilinear grid, which we would not 
have if regions of grid points were missing.

\begin{figure*}[t!]
    \centering
    \includegraphics[width=\textwidth]{images/figure1.png}
    \caption{The number of detected spectral lines at each grid point of 
    a slice of the PHOENIX subset at solar metallicity. We can see that
    the number of detected lines decreases with increasing $T_{\mathrm{eff}}$.
    Also note that from $T_{\mathrm{eff}} = 7000$ K onward, the spacing 
    between grid points increases from 100 K to 200 K.}
\end{figure*}

To solve this, we artificially
populated missing grid points with log-amplitudes of -1000, which retained
interpolator stability by not being an infinity, but also essentially 
nullified the line when evaluated. An example of the appearance of missing 
sections in heatmaps where a line does not appear is shown in Figure 2. 
In total, across the entire PHOENIX subset, \texttt{blase} detected 
128,723 individual spectral lines, meaning we have that many manifolds to 
interpolate.

\begin{figure*}[t!]
    \centering
    \includegraphics[width=\textwidth]{images/figure2.png}
    \caption{Heatmap showing how $\ln(a)$ varies over the PHOENIX subset
    slice at solar metallicity. Notice the missing chunk in the top left
    of the figure; \texttt{blase} did not detect a spectral line here,
    but we have to artificially populate those points with lines that 
    have $\ln(a) = -1000$.}
\end{figure*}

\subsection{Continuously Evaluable Manifolds}
For each line, the inputs to the interpolator were the three input parameters $T_{\mathrm{eff}}$,
$\log(g)$, and [Fe/H], and the output was a list of four parameters, 
$\mu'$, $\ln(a)$, $\sigma$, and $\gamma$. For each line, one of these 
interpolator objects was created using linear interpolation, and these 
interpolators were aggregated into a single list, which was when pickled 
and written to a \texttt{.pkl} file. These interpolators generate multiple 
manifolds with the following mapping:
\begin{gather}
    \mqty[T_{\mathrm{eff}} \\ \log(g) \\ \mathrm{[Fe/H]}] \rightarrow \mqty[\mu' \\ \ln(a) \\ \sigma \\ \gamma]
\end{gather}
These interpolators could now be evaluated at any point lying within the 
domain of the PHOENIX subset, turning a discretely sampled PHOENIX subset 
into a continuous one. With the given size of the PHOENIX subset, the 
interpolator list takes up 13.2 GB of disk space. Since this is now evaluable,
we call this the PHOENIX generator.

\section{Bayesian Inference and Testing}
\subsection{Inference Algorithm}
With the PHOENIX generator complete, we used Bayesian optimization as the
inference algorithm, specifically the \texttt{gp\_minimize} function from
the \texttt{scikit-optimize} library. This algorithm uses a Gaussian Process
to model the objective function, which in this case was the RMS 
(Root-Mean-Square) loss between the interpolated spectrum and the true 
spectrum. The optimizer was configured to first run 30 random evaluations to 
seed the surrogate model, then run 20 more evaluations now guided by the 
surrogate model. This totals to 50 evaluations, which was deemed sufficient 
for this study. We then decided to use ensemble methods, running 32 parallel
optimizers to get a more statistically sound result. One inference run 
takes an average of 15 minutes to complete on Triton (see Section 5 for
specifications).

@@

\section{Discussion}
The computer used for this study, Triton, has the following specifications:
\begin{table}[h!]
    \hspace*{0.6cm}\begin{tabular}{|c|c|}
        \hline
        \multicolumn{2}{|c|}{\textbf{Triton Specs}}\\
        \hline\hline
        CPU & AMD EPYC 7513\\
        Memory & 256 GB\\
        GPU & Nvidia A100 40GB ($\times$2)\\
        \hline
    \end{tabular}
    \caption{This computer is owned by Caroline Morley and is available to
    members of her research group. It was used for all computations, but not 
    for generating visualizations. The EPYC 7513 is a 32c/64t CPU with a 
    boost clock of 3.65 GHz.}
\end{table}
